---
title: "GWAS Diagnostics"
output:
  html_document:
    toc: true
    theme: united
    toc_float: false
params:
  workfolder: unknown
  res1: unknown
  res2: unknown  
  nr_samples_1: 0
  nr_samples_2: 0
  nr_common: 0 
  summaryfile: unknown 
  histplot: unknown
  densityplot: unknown
  qqplot: unknown
  ecdfplot: unknown
  scatterplot: unknown  
---



<!---
https://rmarkdown.rstudio.com/developer_parameterized_reports.html%23parameter_types%2F
The declared parameters are automatically made available within the knit environment as components of a read-only list named params. 
---> 



```{r, echo = FALSE}
workfolder = params$workfolder
res1 = params$res1
res2 = params$res2 
nr_samples_1 = params$nr_samples_1
nr_samples_2 = params$nr_samples_2
nr_common = params$nr_common
summaryfile = params$summaryfile
histplot = params$histplot
densityplot = params$densityplot
qqplot = params$qqplot
ecdfplot = params$ecdfplot
scatterplot = params$scatterplot
```

<br>  



```{r setup, echo = FALSE}
library(knitr)
library(DT, quietly = TRUE)
```







## 1. Overview 


This is a comparison of the residuals stored in <span style="color:red"> `r res1` </span> 
and <span style="color:red"> `r res2` </span>, respectively.

 - The residuals in <span style="color:red"> `r res1` </span> contain `r prettyNum(as.integer(nr_samples_1), big.mark=".", decimal.mark=",")` observations. 
 - The residuals in <span style="color:red"> `r res2` </span> contain `r prettyNum(as.integer(nr_samples_2), big.mark=".", decimal.mark=",")` observations.
 - Both files have `r prettyNum(as.integer(nr_common), big.mark=".", decimal.mark=",")` common observations.



<br> 


## 2. Summary tables  


```{r summary, echo = FALSE, warning = FALSE}
results = get(load(summaryfile))
datatable(results, class = 'cell-border stripe', width = 200, options = list(dom = 't', pageLength = 25), rownames = TRUE)
```

<br> 








## 3. Comparison of the histograms  
 

<br>

```{r, echo = FALSE, fig.cap = "", out.width = '70%', fig.align='left'}
plot = paste(workfolder, histplot, sep ="/") 
knitr::include_graphics(plot)
```

<br> 






## 4. Comparison of the Kernel density plots 


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, densityplot, sep ="/") 
knitr::include_graphics(plot)
```


<br> 




## 5. Comparison of the Quantile-Quantile-plots 


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, qqplot, sep ="/") 
knitr::include_graphics(plot)
```

<br> 







## 6. Comparison of the Empirical Cumulative distribution functions 


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, ecdfplot, sep ="/") 
knitr::include_graphics(plot)
```

<br> 



## 7. Scatterplot residuals 1 vs. residuals 2  


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, scatterplot, sep ="/") 
knitr::include_graphics(plot)
```

<br> 





## 8. Technical information / Help 
 
<br>

 - GWAS workfolder: `r workfolder`
 - <uwe.menzel@medsci.uu.se>   

<br> 











