---
title: "GWAS Report"
output:
  html_document:
    toc: true
    theme: united
    toc_float: false
params:
  workfolder: unknown
  #regtype: unknown
  ident: unknown
  nr_samples: 0
  nr_markers: 0
  nr_sigmarkers: 0
  genoid: unknown
  nr_geno: 0
  lm_call: unknown
  phenoname: unknown 
  plink: unknown
  lambda: unknown 
  mac: unknown  
  maf: unknown 
  vif: unknown 
  sample_max_miss: unknown  
  marker_max_miss: unknown  
  machr2_low: unknown
  machr2_high: unknown   
  hwe_pval: unknown 
  excel_file: unknown
  signif_file: unknown
  filtered_file: unknown
  cojo_done: unknown
  clump_done: unknown  
  cojo_file: unknown 
  cojo_window: unknown  
  cojo_pval: unknown
  cojo_coline: unknown 
  cojo_refgen: unknown  
  cojo_maf: unknown    
  cojo_orig_file: unknown 
  clump_file: unknown
  clump_p1: unknown  
  clump_p2: unknown  
  clump_r2: unknown
  clump_kb: unknown 
  clump_refgen: unknown     
  cojo_neargene_file: unknown
  clump_neargene_file: unknown  
  signif_neargene_file: unknown   
  qq_plot: unknown
  man_plot: unknown
  snp_density_plot: unknown
  histo_plot: unknown
  #cookplot: unknown  
  kernel_plot: unknown
  #datafile: unknown
  #residfile: unknown
  #metricfile: unknown
  #lmfile: unknown
  #viffile: unknown
  #histplot: unknown
  #qqplotres: unknown
  #form: unknown
  #regenie specific params:
  regenie_bsize1: unknown
  regenie_bsize2: unknown
  regenie_trait_type: unknown
  regenie_info: unknown
---




<!-- params = get(load("rmd_review_params.RData")) - saved in review_gwas.R then run code below for testing -->

```{r, echo = FALSE}
workfolder = params$workfolder
#regtype = params$regtype
ident = params$ident
nr_samples = params$nr_samples
nr_markers = params$nr_markers
nr_sigmarkers = params$nr_sigmarkers 
genoid = params$genoid
nr_geno = params$nr_geno
lm_call = params$lm_call
phenoname = params$phenoname
plink = params$plink
lambda = params$lambda
mac = params$mac 
maf = params$maf
vif = params$vif
sample_max_miss = params$sample_max_miss
marker_max_miss = params$marker_max_miss 
machr2_low = params$machr2_low 
machr2_high = params$machr2_high 
hwe_pval = params$hwe_pval  
excel_file = params$excel_file
signif_file = params$signif_file
filtered_file = params$filtered_file
cojo_done = params$cojo_done   	# logical
clump_done = params$clump_done 	# logical
cojo_file = params$cojo_file 
cojo_window = params$cojo_window
cojo_pval = params$cojo_pval
cojo_coline = params$cojo_coline
cojo_refgen = params$cojo_refgen
cojo_maf = params$cojo_maf			
cojo_orig_file = params$cojo_orig_file  	
clump_file = params$clump_file 
clump_p1 = params$clump_p1
clump_p2 = params$clump_p2
clump_r2 = params$clump_r2
clump_kb = params$clump_kb 
clump_refgen = params$clump_refgen		
cojo_neargene_file = params$cojo_neargene_file 	
clump_neargene_file = params$clump_neargene_file
signif_neargene_file = params$signif_neargene_file 
qq_plot = params$qq_plot
man_plot = params$man_plot
snp_density_plot = params$snp_density_plot
histo_plot = params$histo_plot
#cookplot = params$cookplot
kernel_plot = params$kernel_plot 
#datafile = params$datafile
#residfile = params$residfile
#metricfile = params$metricfile
#lmfile = params$lmfile
#viffile = params$viffile
#histplot = params$histplot
#qqplotres = params$qqplotres
regenie_bsize1 = params$regenie_bsize1
regenie_bsize2 = params$regenie_bsize2
regenie_trait_type = params$regenie_trait_type
regenie_info = params$regenie_info
```

<br>  




## Overview


<!-- The overview with the detailed genoid description was saved to ~/bin/overview_with_genoid_description.Rmd -->

These are the GWAS results for job <span style="color:red"> `r ident` </span> 
with `r prettyNum(as.integer(nr_samples), big.mark=".", decimal.mark=",")` samples, 
run for the phenotype <span style="color:red"> `r phenoname` </span>.

<br>


 

## Methods

The association between `r phenoname` and each variant with a minor allele count $\ge$ `r mac` and 
an imputation quality metric $r^2 \ge$ `r machr2_low` was tested using the Regenie software (https://rgcgithub.github.io/regenie/).

A total of `r prettyNum(as.integer(nr_markers), big.mark=".", decimal.mark=",")` markers (after filtering).
Markers having a Hardy-Weinberg equilibrium exact test p-value below `r formatC(hwe_pval, format = "e", digits = 1)` were filtered out.  
`r prettyNum(as.integer(nr_sigmarkers), big.mark=".", decimal.mark=",")` significant markers have been found.
The genomic inflation factor was $\lambda =$ `r lambda`. 


```{r setup, echo = FALSE}
library(knitr)
library(DT, quietly = TRUE)
```

<br>

## GWAS parameters

The regenie program was run with the following parameters:

 - Minor allele count cutoff: `r prettyNum(mac, big.mark=".", decimal.mark=",")`
 - Minimum imputation quality: `r prettyNum(regenie_info, big.mark=".", decimal.mark=",")`
 - Hardy-Weinberg exact test p-value threshold: `r prettyNum(hwe_pval, big.mark=".", decimal.mark=",")`
 - Step 1 genotype block size (bsize): `r prettyNum(regenie_bsize1, big.mark=".", decimal.mark=",")`
 - Step 2 genotype block size (bsize): `r prettyNum(regenie_bsize2, big.mark=".", decimal.mark=",")`
 - Trait type: `r prettyNum(regenie_trait_type, big.mark=".", decimal.mark=",")`
 
<br>




## Filtered samples and variants

The table displays the number of samples/variants that have been removed due to different filters:

 - **N_SAMPLES**: Samples loaded from thte genotype file.
 - **N_SNPS**: Number of SNPs loaded from the genotype file.
 - **PHENO_SAMPLES_NOMISS**: Number of samples with phenotype values.
 - **COVAR_SAMPLES**:  Number of samples with covariate values.
 - **ANALYSIS_SAMPLES**: Number of samples used in the analysis.
 - **SKIPPED_MAC_INFO**:  Number of variants skipped due to MAC and INFO score filters.

  
```{r filter, echo = FALSE, warning = FALSE}
filtered_file = paste(workfolder, filtered_file, sep ="/")  # "/castor/project/proj/GWAS_TEST/LIV_MULT4/LIV_MULT4_liv2_cojoed_markers.RData"
if(file.exists(filtered_file)) {
  filtered = get(load(filtered_file))
} else {
  stop(paste("\n\n  ERROR (gwas_report.Rmd): File", filtered_file, "not found.\n\n"))
}
```

<!-- need extra chunk for data table -->

```{r filter_table, echo = FALSE, warning = FALSE}
options(htmlwidgets.TOJSON_ARGS = list(na = 'string'))   # show NA (no empty fields!) 
datatable(filtered, class = 'cell-border stripe', rownames = FALSE, width = 600, options = list(dom = 'tp')) 
```

<br>




<!-- Table with genome-wide significant markers --> 

```{asis, echo = signif_file == "none"}
## Genome-wide significant markers
No genome-wide significant markers found.
```

```{r, child = if (signif_file != "none") 'table_signif.Rmd'}
```

<br>



<!-- Table with GCTA-COJO results -->


```{asis, echo = !cojo_done}
## GCTA-COJO results
GCTA cojo was not conducted.
```

```{asis, echo = cojo_done & cojo_file == "none"}
## GCTA-COJO results
No independent markers identified by GCTA cojo.
```

```{r, child = if (cojo_file != "none") 'table_cojoed.Rmd'}
```

<br>


<!-- Table with plink-clump results -->

```{asis, echo = !clump_done}
## CLUMP results
Plink clumping was not conducted.
```

```{asis, echo = clump_done & clump_file == "none"}
## CLUMP results
No independent markers identified by plink clump.
```

```{r, child = if (clump_file != "none") 'table_clumped.Rmd'}
```

<br>




<!-- Compare cojo with plink-clump results -->


```{r, child = if (clump_file != "none" & cojo_file != "none") 'comparison_plink_clump.Rmd'}
```

<br> 



<!-- Show all significant markers: only if neither cojo nor clump was conducted, and if significant markers were found --> 


```{r, child = if (clump_file == "none" & cojo_file == "none" & signif_file != "none") 'table_unpruned.Rmd'}
```
<br> 



## QQ-plot for $p$-values

```{r, echo=FALSE, fig.cap="", out.width = '100%', fig.align='left'}
plot = paste(workfolder, qq_plot, sep ="/")  
knitr::include_graphics(plot)
```


## Manhattan plot

```{r, echo=FALSE, fig.cap="", out.width = '100%', fig.align='left'}
plot = paste(workfolder, man_plot, sep ="/") 
knitr::include_graphics(plot)
```


## SNP density plot

```{r, echo=FALSE, fig.cap="", out.width = '100%', fig.align='left'}
plot = paste(workfolder, snp_density_plot, sep ="/") 
knitr::include_graphics(plot)
```



## Histogram of $\beta$ values 

```{r, echo=FALSE, fig.cap="", out.width = '100%', fig.align='left'}
plot = paste(workfolder, histo_plot, sep ="/")
knitr::include_graphics(plot)
```


## Kernel density plot of $\beta$ values 

```{r, echo=FALSE, fig.cap="", out.width = '100%', fig.align='left'}
plot = paste(workfolder, kernel_plot, sep ="/")
knitr::include_graphics(plot)
```

<br> 



## Technical information / Help   


```{r params, echo = FALSE, warning = FALSE}   
parfile = paste(ident, "_gwas_params.txt", sep="")
parfile_path = paste(workfolder, parfile, sep = "/") 
```

 - GWAS **workfolder** `r workfolder`
 - **Parameter file**: `r parfile_path`   
 - <uwe.menzel@medsci.uu.se> 
 
<br> 


















