---
title: "Linkage between two markers"
output:
  html_document:
    toc: true
    theme: united
    toc_float: false
params:
  workfolder: unknown
  genoid: unknown
  rawfile1: unknown
  rawfile2: unknown
  snp1: unknown
  snp2: unknown
  nr_obs1: 0
  nr_obs2: 0
  imput1: unknown
  imput2: unknown
  nr_imputed1: 0
  nr_imputed2: 0  
  histplot_imp1: unknown
  histplot_imp2: unknown
  histplot1: unknown
  histplot2: unknown
  form: unknown
  metricfile: unknown
  scatterplot: unknown
  histplotres: unknown
  qqplot: unknown
---



```{r, echo = FALSE}
workfolder = params$workfolder
genoid = params$genoid
rawfile1 = params$rawfile1     
rawfile2 = params$rawfile2     
snp1 = params$snp1	
snp2 = params$snp2	
nr_obs1 = params$nr_obs1	
nr_obs2 = params$nr_obs2	
imput1 = params$imput1	
imput2 = params$imput2
nr_imputed1 = params$nr_imputed1
nr_imputed2 = params$nr_imputed2	
histplot_imp1 = params$histplot_imp1
histplot_imp2 = params$histplot_imp2
histplot1 = params$histplot1
histplot2 = params$histplot2
form = params$form
metricfile = params$metricfile
scatterplot = params$scatterplot
histplotres = params$histplotres
qqplot = params$qqplot	
```

<br>  



```{r setup, echo = FALSE}
library(knitr)
library(DT, quietly = TRUE)
```




## Linear Regression 


The linear regression model was:  

<span style="color:blue">
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;  `r form` 
</span>

The samples originate from the genotype dataset 
with identifier <span style="color:blue">`r genoid`</span>, comprising `r prettyNum(nr_obs1, big.mark=".", decimal.mark=",")` observations.
The extracted genotype datasets were stored in *`r rawfile1`* and *`r rawfile2`*, respectively.




<br> 

## Genotype vs. genotype plot


```{r imputation, echo = FALSE, results = 'asis', warning = FALSE}
if(imput1) {
  cat(paste("  The marker", snp1, "contains", nr_imputed1, "imputed values.\n"))
} else {
  cat(paste(" Note that", snp1, "does not contain imputed values!\n"))   
}

if(imput2) {
  cat(paste("  The marker", snp2, "contains", nr_imputed2, "imputed values.\n"))
} else {
  cat(paste(" Note that", snp2, "does not contain imputed values!\n"))    
}
```


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, scatterplot, sep ="/") 
knitr::include_graphics(plot)
```


<br> 




## Some metrics obtained using `lm()`:

- **sigma** is the estimatimated standard deviation of the noise term 
- **Fstat** is the value of the F-statistic
- **Rsquared** is coefficient of determination 
- **Rsq.adj** is coefficient of determination adjusted for the number of predictors
- **$\beta$** is the slope of the regression line
- **CI_low** is the lower limit of the confidence interval for $\beta$
- **CI_up** is the upper limit of the confidence interval for $\beta$

```{r metrics, echo = FALSE, warning = FALSE} 
# options(scipen = 999)
results = get(load(metricfile))
results = signif(results,3)  
datatable(t(results), class = 'cell-border stripe', options = list(dom = 't'), rownames = FALSE) %>% formatStyle('Rsquared', color = 'green') 
```

<br> 




##  Histogram for the genotype values


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, histplot1, sep ="/") 
knitr::include_graphics(plot)
```

<br> 


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, histplot2, sep ="/") 
knitr::include_graphics(plot)
```

<br> 





##  Histogram for the imputed genotype values

Note that some markers do not comprise imputed values.

<br>

```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left', eval = imput1}
plot = paste(workfolder, histplot_imp1, sep ="/") 
knitr::include_graphics(plot)
```

<br> 


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left', eval = imput2}
plot = paste(workfolder, histplot_imp2, sep ="/") 
knitr::include_graphics(plot)
```

<br> 





##  Histogram for the magnitudes of the residuals

According to the linear model established, the residuals should be normally distributed. 
Consequently, the histogram below should approximately resemble a normal distribution.


```{r, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, histplotres, sep ="/") 
knitr::include_graphics(plot)
```

<br> 






## Normal Q-Q plot of the residuals

This plot shows if residuals are normally distributed. 
It is desirable that the points displaying the residuals are located close to the straight line.


```{r qqplot, echo=FALSE, fig.cap="", out.width = '70%', fig.align='left'}
plot = paste(workfolder, qqplot, sep ="/") 
knitr::include_graphics(plot)
```

<br>  





## Technical information / Help  
 
<br>

 - GWAS workfolder: `r workfolder`
 - <uwe.menzel@medsci.uu.se>  
 
<br> 

